### 2012-04-06 mindreading/minimal theory of mind (new version)/minimal theory of mind.nv.tex
mtom revision sent to Ian

\begin{abstract}
\noindent
What could someone represent that would enable her to track, at least within limits, perceptions, knowledge states and beliefs including false beliefs? 
An obvious possibility is that she might represent these very propositional attitudes as such.
It is sometimes tacitly or explicitly assumed that this is the only possible answer.
However we argue that several recent discoveries in developmental, cognitive, and comparative psychology indicate the need for other, less obvious possibilities.
Our aim is to meet this need by describing the construction of a minimal theory of mind.  
Minimal theory of mind is rich enough to explain systematic success on tasks held to be acid tests for theory of mind cognition including many false belief tasks.  
It is also extensible in ways that can explain a wide range of findings from non-human animals and human infants that are sometimes presented as evidence for full-blown theory of mind cognition.  
Yet minimal theory of mind does not require representing propositional attitudes, or any other kind of representation, as such.
Minimal theory of mind may be what enables those with limited cognitive resources or little conceptual sophistication, such as infants, chimpanzees, scrub-jays and human adults under load, able to track, within limits, facts about perceptions and beliefs.

@@

\section{Introduction}
What could someone represent that would enable her to track, at least within limits, perceptions, knowledge states and beliefs including false beliefs? 
One answer is obvious: she might track these things by virtue of representing them as such, that is, by representing perceptions, beliefs, and other propositional attitudes as such.
Our aim in what follows is to identify another, less obvious answer.
There is a form of cognition---minimal theory of mind---which does not involve representing propositional attitudes as such but does involve representing simpler, relational mental states which could, within limits, enable one to track propositional attitudes such as beliefs.
Minimal theory of mind 
is  rich enough to enable systematic success on tasks held to be acid tests for theory of mind cognition including many false belief tasks.
As we will explain, this has consequences 
for interpreting a range of findings concerning infants', adults' and nonhumans' performances on theory of mind tasks.
It may help us to understand what enables those with limited cognitive resources or little conceptual sophistication, such as infants, chimpanzees, scrub-jays and human adults under load, to track, within limits, facts about perceptions and beliefs.

@@

Theory of mind abilities are not only found in humans.
For instance,
scrub-jays can selectively re-cache their food in ways that deprive competitors of knowledge of its location \citep{Clayton:2007fh}, and  chimpanzees can both select routes to approach food which conceal them from a competitor’s view \citep[]{en_1546} and also retrieve food using strategies that optimize their return given what a dominant competitor has seen \citep[]{en_1545}.  
There is debate about the cognitive underpinnings of these abilities. 
Some researchers argue that they may involve representations of nonintentional behaviour only,\footnote{
For example, on chimpanzees: 
	\citet{povinelli:2004vonk}; 
	\citet[pp.\ 364-5]{Vonk:2006cq}; and on scrub-jays and chimpanzees \citet{Penn:2007ey}.
} 
others that they involve representations of perceptions, knowledge states and other propositional attitudes.\footnote{
For example, on chimpanzees: \citet{Tomasello:2005ce,
			Call:2008di}; and on scrub-jays: 
			\citet[p.\ 73]{Emery:2007ze}.
}
If these arguments are not yet decisive
and if the available evidence does not already tightly constrain the space of admissible conjectures,
then the construction of a minimal theory of mind may also be relevant to these debates.
For the conjecture that minimal theory of mind explains chimpanzees' or scrub-jays' abilities to track perceptions or beliefs can be empirically distinguished from conjectures about representations of nonintentional behaviour only and from conjectures about representations of perceptions, knowledge states and beliefs (as we explain in section *** below).

@@

\section{Minimal theory of mind}

In this section we begin with someone, call her Lucky, capable  of representing nonintentional behaviour only and ask what more is needed for minimal theory of mind cognition.  We describe Lucky’s progress with a series of principles. The principles are constructed in such a way that it would be coherent to suppose that Lucky has the abilities codified by the first \textit{n} principles only. They are not intended to represent a developmental or evolutionary progression.  The principles can also be extended to explain more sophisticated theory of mind abilities than those considered here.  We restrict ourselves to these five principles because they are sufficient to explain success on some false belief tasks.%
\footnote{
In standard false belief tasks, `[t]he subject is aware that he/she and another person [call him Maxi] witness a certain state of affairs x.  Then, in the absence of the other person the subject witnesses an unexpected change in the state of affairs from x to y' \citep[][p.\ 106]{en_89}.  The test concerns whether the subject realises that Maxi will falsely believe x to obtain.  In many cases the states of affairs, x and y, differ only with respect to the location of an object \citep[e.g.][]{en_1092, en_1208, en_1824}. 
As we go on to discuss, our proposal for a minimal theory of mind could be extended to cover a range of other cases;
but importantly there are also false belief tasks success on which cannot be explained by minimal theory of mind cognition (see section ***).
}

@@

Of course it was never in question that adults represent encounters or perceptions and not only behaviours; the relevance of these studies is that they demonstrate that altercentric interference can be used as an indirect measure for distinguishing behavioural strategies from minimal theory of mind cognition.  \citet[]{kovacs_social_2010} report evidence of altercentric interference in infants’ expectations about the presence of an object.  Their study suggests that infants’ surprise when their expectations about an object’s presence or absence are violated is modulated by facts about another’s expectations about that object’s location.  Infants looked for less time at the unexpected presence of an object behind a screen when another observer incorrectly expected it to be there than when the observed did not expect it to be there. These findings seem to be evidence of altercentric interference, which is consistent with either minimal or full-blown theory of mind but difficult to explain if participants are exclusively using behavioural strategies.

@@

To answer this question we constructed a minimal theory of mind.  The construction is sufficiently rich to explain systematic success on tasks held to be acid tests for theory of mind cognition including many false belief tasks.  It is also extensible in ways that can explain a range of findings from non-human animals and human infants that are often presented as evidence for full-blown theory of mind cognition.  Where minimal theory of mind must break down is in cases involving quantification or mistakes about identity.  Because such cases require full-blown theory of mind, it is possible to distinguish whether an individual’s performance on a particular task involves minimal or full-blown theory of mind cognition.

The central notions for minimal theory of mind are encountering and registration.  We characterised these by their structures, acquisition conditions and roles in producing goal-directed actions.  Encountering and registration serve as non-representational proxies for perception and belief in this sense: in a limited but useful range of everyday circumstances, agents perceive an object just when they encounter it and they believe that an object has a given property just when they register it having that property.  But encountering differs from perception in not constitutively involving appearances, modalities or the possibility of illusion and in not being constitutively linked to reasons, knowledge or informational states.  Similarly, registration differs from belief in lacking representational content and the possibility of embedding, in not interacting in arbitrarily complex ways with other states, and in the simplicity of its parameter-setting role in causing actions.  These differences ensure that representing encountering or registration need not involve meeting conceptual and cognitive demands associated with representing perceptions, beliefs and other propositional attitudes as such.

The novelty of our constructive approach lies in several features.  It does not rely directly on everyday psychological concepts, whose exact nature is a source of controversy.
Nor does it rely on infants or non-human animals holding theoretical commitment to simplified versions of these concepts (contrast 
	\citealp{en_75};
	\citealp{en_81};
	\citealp{en_2372} and
	\citealp{en_1614}%
).
Importantly, we do not assume that minimal theory of mind develops into full-blown theory of mind in humans.  
It may instead remain distinct, supporting cognitively efficient theory of mind across the lifespan \citep[see][]{Samson:2010jm, en_2397}.  The construction of minimal theory of mind is systematic enough to generate testable predictions distinguishing it from both behavioural stratetgies and full-blown theory of mind cognition (see Sections \ref{sec:limits1} and \ref{sec:limits2}).  
Our construction makes detailed sense of the notion that there are degrees of theory of mind cognition.  
(This is a virtue because while is widely recognised that degrees of theory of mind cognition are needed \citep[e.g.][]{en_610, en_1415}, there have been few detailed attempts to make systematic sense of this possibility.)
It also pushes much further than earlier work the boundaries of what can be achieved without full-blown theory of mind cognition; in particular, it explains how systematic success on a range of false belief tasks (but not those which essentially involve identity or quantification) is possible without representing beliefs or other propositional attitudes as such. Minimal theory of mind may be what enables those with limited cognitive resources or little conceptual sophistication, such as infants, chimpanzees, scrub-jays and human adults under load,  to track  perceptions, knowledge states and beliefs.

>>>>

\begin{abstract}
\noindent
What could someone represent that would enable her to track, at least within limits, others' perceptions, knowledge states and beliefs including false beliefs? 
An obvious possibility is that she might represent these very attitudes as such.
It is sometimes tacitly or explicitly assumed that this is the only possible answer.
However we argue that several recent discoveries in developmental, cognitive, and comparative psychology indicate the need for other, less obvious possibilities.
Our aim is to meet this need by describing the construction of a minimal theory of mind.  
Minimal theory of mind is rich enough to explain systematic success on tasks held to be acid tests for theory of mind cognition including many false belief tasks.  
Yet minimal theory of mind does not require representing propositional attitudes, or any other kind of representation, as such.
Minimal theory of mind may be what enables those with limited cognitive resources or little conceptual sophistication, such as infants, chimpanzees, scrub-jays and human adults under load,  to track others' perceptions, knowledge states and beliefs.

@@

\section{Introduction}
What could someone represent that would enable her to track, at least within limits, others' perceptions, knowledge states and beliefs including false beliefs? 
One answer is obvious: she might track these things by virtue of representing them as such, that is, by representing perceptions, beliefs, and other propositional attitudes as such.
Our aim in what follows is to identify another, less obvious answer.
There is a form of cognition---minimal theory of mind---which does not involve representing propositional attitudes as such but does involve representing simpler, relational mental states which could, within limits, enable one to track propositional attitudes such as beliefs.
Minimal theory of mind 
is  rich enough to enable systematic success on tasks held to be acid tests for theory of mind cognition including many false belief tasks.
As we will explain, this has consequences 
for interpreting a range of findings concerning infants', adults' and nonhumans' performances on theory of mind tasks.
It may help us to understand what enables those with limited cognitive resources or little conceptual sophistication, such as infants, chimpanzees, scrub-jays and human adults under load, to track, within limits, facts about perceptions and beliefs.

@@

Theory of mind abilities are not only found in humans.
For instance,
scrub-jays can selectively re-cache their food in ways that deprive competitors of knowledge of its location \citep{Clayton:2007fh}, and  chimpanzees can both select routes to approach food which conceal them from a competitor’s view \citep[]{en_1546} and also retrieve food using strategies that optimize their return given what a dominant competitor has seen \citep[]{en_1545}.  
There is debate about the cognitive underpinnings of these abilities. 
Some researchers argue that they may involve representations of nonintentional behaviour only,\footnote{
For example, on chimpanzees: 
	\citet{povinelli:2004vonk}; 
	\citet[pp.\ 364-5]{Vonk:2006cq}; and on scrub-jays and chimpanzees \citet{Penn:2007ey}.
} 
others that they involve representations of perceptions, knowledge states and other propositional attitudes.\footnote{
For example, on chimpanzees: \citet{Tomasello:2005ce,
			Call:2008di}; and on scrub-jays: 
			\citet[p.\ 73]{Emery:2007ze}.
}
If these arguments are not yet decisive
and if the available evidence does not already tightly constrain the space of admissible hypotheses,
then the construction of a minimal theory of mind may also be relevant to these debates.
For the hypothesis that minimal theory of mind explains particular subjects' abilities to track perceptions or beliefs can be empirically distinguished from hypotheses about representations of nonintentional behaviour only and also from hypotheses about representations of perceptions, knowledge states and beliefs (as we explain in Sections \ref{sec:limits1} and \ref{sec:limits2} below).

@@

\section{Minimal theory of mind}
\label{sec:minimal}
In this section we begin with someone, call her Lucky, capable  of representing nonintentional behaviour only and ask what more is needed for minimal theory of mind cognition.  We describe Lucky’s progress with a series of principles. The principles are constructed in such a way that it would be coherent to suppose that Lucky has the abilities codified by the first \textit{n} principles only. They are not intended to represent a developmental or evolutionary progression.  The principles can also be extended to explain more sophisticated theory of mind abilities than those considered here.  We restrict ourselves to these five principles because they are sufficient to explain success on some false belief tasks.%
\footnote{
In standard false belief tasks, `[t]he subject is aware that he/she and another person [call him Maxi] witness a certain state of affairs x.  Then, in the absence of the other person the subject witnesses an unexpected change in the state of affairs from x to y' \citep[][p.\ 106]{en_89}.  The test concerns whether the subject realises that Maxi will falsely believe x to obtain.  In many cases the states of affairs, x and y, differ only with respect to the location of an object \citep[e.g.][]{en_1092, en_1208, en_1824}. 
As we go on to discuss, our proposal for a minimal theory of mind could be extended to cover a range of other cases;
but importantly there are also false belief tasks success on which cannot be explained by minimal theory of mind cognition (see Section \vref{sec:limits1}).
}

@@

Of course it was never in question that adults represent encounters or perceptions and not only behaviours; the relevance of these studies is that they demonstrate that altercentric interference can be used as an indirect measure for distinguishing behavioural strategies from minimal theory of mind cognition.  \citet[]{kovacs_social_2010} report evidence of possible altercentric interference in infants’ expectations about the presence of an object.  Their study suggests that infants’ surprise when their expectations about an object’s presence or absence are violated is modulated by facts about another’s expectations about that object’s location.  Infants looked for less time at the unexpected presence of an object behind a screen when another observer incorrectly expected it to be there than when the observed did not expect it to be there. These findings may be evidence of altercentric interference,%
\footnote{
An anonymous referee pointed out that  this is not the only possible interpretation.
% raised the possibility that the pattern of infants'  looking times in this experiment might be explained not by their representing others' expectations but rather by their representing others' presence or absence at critical moments.  
We do not mean to suggest that a single study provides decisive evidence for altercentric interference, only that in the wider context of research on infant theory of mind  it provides an indication and constitutes progress on the question of how altercentric interference could be measured.
}
which would be consistent with either minimal or full-blown theory of mind but difficult to explain in a principled way if subjects are  using behavioural strategies only.

@@

To answer this question we constructed a minimal theory of mind.  The construction is  rich enough to explain systematic success on tasks held to be acid tests for theory of mind cognition including many false belief tasks.  
%It is also extensible in ways that can explain a range of findings from non-human animals and human infants that are often presented as evidence for full-blown theory of mind cognition.  
Where minimal theory of mind must break down is in cases involving quantification or mistakes about identity (see Section \ref{sec:limits1}).  
Because such cases require full-blown theory of mind, it is possible to distinguish whether an individual’s performance on a particular task involves minimal or full-blown theory of mind cognition.

The central notions for minimal theory of mind are encountering and registration.  We characterised these by their structures, acquisition conditions and roles in producing goal-directed actions (see Section \ref{sec:minimal}).  Encountering and registration serve as non-representational proxies for perception and belief in this sense: in a limited but useful range of everyday circumstances, agents perceive an object just when they encounter it and they believe that an object has a given property just when they register it having that property.  But encountering differs from perception in not constitutively involving appearances, modalities or the possibility of illusion and in not being constitutively linked to reasons, knowledge or informational states.  Similarly, registration differs from belief in lacking representational content and the possibility of embedding, in not interacting in arbitrarily complex ways with other states, and in the simplicity of its parameter-setting role in causing actions.  These differences ensure that representing encountering or registration need not involve meeting conceptual and cognitive demands associated with representing perceptions, beliefs and other propositional attitudes as such.

The novelty of our constructive approach lies in several features.  It does not rely directly on everyday psychological concepts, whose exact nature is a source of controversy.
Nor does it rely on infants or non-human animals holding theoretical commitment to simplified versions of these concepts (contrast 
	\citealp{en_75};
	\citealp{en_81};
	\citealp{en_2372} and
	\citealp{en_1614}%
).
Importantly, we do not assume that minimal theory of mind develops into full-blown theory of mind in humans.  
It may instead remain distinct, supporting cognitively efficient theory of mind across the lifespan \citep[see][]{Samson:2010jm, en_2397}.  The construction of minimal theory of mind is systematic enough to generate testable predictions distinguishing it from both behavioural stratetgies and full-blown theory of mind cognition (see Sections \ref{sec:limits1} and \ref{sec:limits2}).  
Our construction makes detailed sense of the notion that there are degrees of theory of mind cognition.  
(This is a virtue because while is widely recognised that degrees of theory of mind cognition are needed \citep[e.g.][]{en_610, en_1415}, there have been few detailed attempts to make systematic sense of this possibility.)
It also pushes much further than earlier work the boundaries of what can be achieved without full-blown theory of mind cognition; in particular, it explains how systematic success on a range of false belief tasks (but not those which essentially involve identity or quantification) is possible without representing beliefs or other propositional attitudes as such. Minimal theory of mind may be what enables those with limited cognitive resources or little conceptual sophistication, such as infants, chimpanzees, scrub-jays and human adults under load,  to track others'  perceptions, knowledge states and beliefs.
